plyranges: a grammar for manipulating genomics data

### 2018-02-12 paper.Rmd
some analysis interspersed in the text

bibliography: references.bib
output: rticles::plos_article
csl: plos.csl
---

# Introduction

@@

@@

The _plyranges_ API provides the methods `as_granges()` and `as_iranges()`
for constructing Ranges from tabular data structures, such as data.frames in
base R. These methods use non-standard evaluation so columns in a data.frame
can be modified before a Ranges object is created.

The package also provides a consistent framework for reading and writing 
files from and to common genomic data formats, using the `rtracklayer` package
as a backend. The methods are implemented in the `read_/write_` family of 
functions, currently _plyranges_ can read and write BAM, BED, BEDPE, 
narrowPeaks, GFF/GTF, WIG and BigWig files.

@@

@@

A common operation to perform between two _Ranges_ objects is to find overlaps
or nearest neighbours. The _plyranges_ API recasts these operations as 'joins'
or 'pairing' operations. For overlaps, there are three join operations: 
`join_overlap_intersect()`, `join_overlap_inner()` and `join_overlap_left()`
which are shown in figure (). These operations consider any overlap between
two input ranges and return any corresponding metadata from both Ranges objects
as metadata. The intersect join takes the intersect of the start and end coordinates
of overlapping intervals of the query and subject Ranges 
(for GenomicRanges it also accounts for sequence name), when there is a overlap
the metadata corresponding to the query and subject Ranges are returned. 
Similarly, inner join takes the start and end coordinates of the query Ranges
that overlap the subject Ranges and returns metadata of the overlapping query
and subject Ranges. Finally, the left join performs a left outer join between
the query and subject Ranges, it returns all genomic intervals from the query
ranges, and returns missing values in metadata columns when there is no overlap.
A user may also restrict or group by overlaps with the `filter_by_overlaps()`,
`filter_by_non_overlaps()` and `group_by_overlaps()`. All overlap methods can
be modified with the `within` suffix (which changes the type of overlap 
from 'any' to 'within') or the `directed` suffix (which takes into account
the strand of a GenomicRanges object.).

@@

### Non-standard evaluation

>>>>

bibliography: references.bib
output: rticles::plos_article
csl: plos.csl
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, comment = "#>", 
                      fig.width = 6, fig.height = 4)
library(ggbio)
library(ggplot2)
library(plyranges)
```
# Introduction

@@

In order to provide a compatible API with _dplyr_, _plyranges_ makes use
of non-standard evaluation in R via the _rlang_ package. 
<!--- need to mention something about how this could be detrimental --->

@@

The _plyranges_ API provides the methods `as_granges()` and `as_iranges()`
for constructing Ranges from tabular data structures, such as data.frames in
base R. These methods use non-standard evaluation so columns in a data.frame
can be modified before a Ranges object is created. The API also has
convienence methods for annotating or extracting annotations from Genomic Ranges
objects with the `set_genome_info()` and `get_genome_info()` functions.

There is a consistent framework for reading and writing 
files from and to common genomic data formats, using the `rtracklayer` package
as a backend. The methods are implemented in the `read_/write_` family of 
functions, currently _plyranges_ can read and write BAM, BED, BEDPE, 
narrowPeaks, GFF/GTF, WIG and BigWig files.

As an example, we use the Bioconductor package AnnotationHub to search for 
BigWig files from ChIP-Seq experiments from the Human Epigenome Roadmap project.
We choose to focus on assays for primary T CD8+ memory cells from peripheral 
blood. We can then read the BigWig file  corresponding to the H3K27Me3 
methylation mark over chromosome 10. First we gather the BigWig file
and extract its annotation information and filter it to chromosome 10. 
Then we read the BigWig file only extracting scores if they overlap chromosome
10. We then add annotation information to ensure we know we are working
with the hg19 genome build.

```{r load-bw, cache = TRUE}
# E048 corresponds to CD8+ T cells from peripheral blood
library(plyranges)
library(AnnotationHub)
q <-query(AnnotationHub(), c("EpigenomeRoadMap", "E048"))

bw_file <- q[["AH33458"]]

chr10_ranges <- bw_file %>% 
  get_genome_info() %>%
  filter(seqnames == "chr10")

chr10_scores <- bw_file %>%
  read_bigwig(overlap_ranges = chr10_ranges) %>%
  set_genome_info(genome = "hg19")
chr10_scores
```

@@

Returning to the Ranges object containing normalised coverage scores for
the methylation data, we can filter to find the coordinates of the peak 
containing maximum coverage score. We can then find 
a 5000 nt region centered around the maximum position.

```{r max-score-region}
chr10_max_score_region <- chr10_scores %>%
  filter(score == max(score)) %>% 
  anchor_center() %>%
  set_width(5000)
```

@@

A common operation to perform between two _Ranges_ objects is to find overlaps
or nearest neighbours. The _plyranges_ API recasts these operations as 'joins'
or 'pairing' operations. For overlaps, there are three join operations: 
`join_overlap_intersect()`, `join_overlap_inner()` and `join_overlap_left()`
which are shown in figure (). These operations consider any overlap between
two input ranges and return any corresponding metadata from both Ranges objects
as metadata. The intersect join takes the intersect of the start and end coordinates
of overlapping intervals of the query and subject Ranges 
(for GenomicRanges it also accounts for sequence name), when there is a overlap
the metadata corresponding to the query and subject Ranges are returned. 
Similarly, inner join takes the start and end coordinates of the query Ranges
that overlap the subject Ranges and returns metadata of the overlapping query
and subject Ranges. Finally, the left join performs a left outer join between
the query and subject Ranges, it returns all genomic intervals from the query
ranges, and returns missing values in metadata columns when there is no overlap.

The overlap innner join could be used to restrict the chromosome 10
normalised coverage scores that are within the 5000nt region that contains
the max peak on chromosome 10 (visualised in figure ()).

```{r peak_region, fig.align = "center", fig.cap = "Visualisation of peak coverage score accross 5000nt region of chromosome 10"}
peak_region <- chr10_scores %>%
  join_overlap_inner(., 
                     chr10_max_score_region %>% 
                       select(-score))
ggbio::autoplot(peak_region, 
                aes(y = score), 
                geom = "blank") + 
  geom_area(aes(x = start))
```

A user may also restrict or group by overlaps with the `filter_by_overlaps()`,
`filter_by_non_overlaps()` and `group_by_overlaps()`. All overlap methods can
be modified with the `within` suffix (which changes the type of overlap 
from 'any' to 'within') or the `directed` suffix (which takes into account
the strand of a GenomicRanges object.).

@@

### 2018-02-12 paper.tex
some analysis interspersed in the text

@@

@@

The \emph{plyranges} API provides the methods \texttt{as\_granges()} and
\texttt{as\_iranges()} for constructing Ranges from tabular data
structures, such as data.frames in base R. These methods use
non-standard evaluation so columns in a data.frame can be modified
before a Ranges object is created.

The package also provides a consistent framework for reading and writing
files from and to common genomic data formats, using the
\texttt{rtracklayer} package as a backend. The methods are implemented
in the \texttt{read\_/write\_} family of functions, currently
\emph{plyranges} can read and write BAM, BED, BEDPE, narrowPeaks,
GFF/GTF, WIG and BigWig files.

@@

@@

A common operation to perform between two \emph{Ranges} objects is to
find overlaps or nearest neighbours. The \emph{plyranges} API recasts
these operations as `joins' or `pairing' operations. For overlaps, there
are three join operations: \texttt{join\_overlap\_intersect()},
\texttt{join\_overlap\_inner()} and \texttt{join\_overlap\_left()} which
are shown in figure (). These operations consider any overlap between
two input ranges and return any corresponding metadata from both Ranges
objects as metadata. The intersect join takes the intersect of the start
and end coordinates of overlapping intervals of the query and subject
Ranges (for GenomicRanges it also accounts for sequence name), when
there is a overlap the metadata corresponding to the query and subject
Ranges are returned. Similarly, inner join takes the start and end
coordinates of the query Ranges that overlap the subject Ranges and
returns metadata of the overlapping query and subject Ranges. Finally,
the left join performs a left outer join between the query and subject
Ranges, it returns all genomic intervals from the query ranges, and
returns missing values in metadata columns when there is no overlap. A
user may also restrict or group by overlaps with the
\texttt{filter\_by\_overlaps()}, \texttt{filter\_by\_non\_overlaps()}
and \texttt{group\_by\_overlaps()}. All overlap methods can be modified
with the \texttt{within} suffix (which changes the type of overlap from
`any' to `within') or the \texttt{directed} suffix (which takes into
account the strand of a GenomicRanges object.).

@@

\hypertarget{non-standard-evaluation}{%
\subsubsection{Non-standard evaluation}\label{non-standard-evaluation}}

>>>>

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

@@

In order to provide a compatible API with \emph{dplyr}, \emph{plyranges}
makes use of non-standard evaluation in R via the \emph{rlang} package.

@@

The \emph{plyranges} API provides the methods \texttt{as\_granges()} and
\texttt{as\_iranges()} for constructing Ranges from tabular data
structures, such as data.frames in base R. These methods use
non-standard evaluation so columns in a data.frame can be modified
before a Ranges object is created. The API also has convienence methods
for annotating or extracting annotations from Genomic Ranges objects
with the \texttt{set\_genome\_info()} and \texttt{get\_genome\_info()}
functions.

There is a consistent framework for reading and writing files from and
to common genomic data formats, using the \texttt{rtracklayer} package
as a backend. The methods are implemented in the \texttt{read\_/write\_}
family of functions, currently \emph{plyranges} can read and write BAM,
BED, BEDPE, narrowPeaks, GFF/GTF, WIG and BigWig files.

As an example, we use the Bioconductor package AnnotationHub to search
for BigWig files from ChIP-Seq experiments from the Human Epigenome
Roadmap project. We choose to focus on assays for primary T CD8+ memory
cells from peripheral blood. We can then read the BigWig file
corresponding to the H3K27Me3 methylation mark over chromosome 10. First
we gather the BigWig file and extract its annotation information and
filter it to chromosome 10. Then we read the BigWig file only extracting
scores if they overlap chromosome 10. We then add annotation information
to ensure we know we are working with the hg19 genome build.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# E048 corresponds to CD8+ T cells from peripheral blood}
\KeywordTok{library}\NormalTok{(plyranges)}
\KeywordTok{library}\NormalTok{(AnnotationHub)}
\NormalTok{q <-}\KeywordTok{query}\NormalTok{(}\KeywordTok{AnnotationHub}\NormalTok{(), }\KeywordTok{c}\NormalTok{(}\StringTok{"EpigenomeRoadMap"}\NormalTok{, }\StringTok{"E048"}\NormalTok{))}

\NormalTok{bw_file <-}\StringTok{ }\NormalTok{q[[}\StringTok{"AH33458"}\NormalTok{]] }

\NormalTok{chr10_ranges <-}\StringTok{ }\NormalTok{bw_file }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{get_genome_info}\NormalTok{() }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{filter}\NormalTok{(seqnames }\OperatorTok{==}\StringTok{ "chr10"}\NormalTok{)}

\NormalTok{chr10_scores <-}\StringTok{ }\NormalTok{bw_file }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{read_bigwig}\NormalTok{(}\DataTypeTok{overlap_ranges =}\NormalTok{ chr10_ranges) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{set_genome_info}\NormalTok{(}\DataTypeTok{genome =} \StringTok{"hg19"}\NormalTok{)}
\NormalTok{chr10_scores}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#> GRanges object with 5789841 ranges and 1 metadata column:
#>             seqnames                 ranges strand |              score
#>                <Rle>              <IRanges>  <Rle> |          <numeric>
#>         [1]    chr10         [    1, 60602]      * | 0.0422799997031689
#>         [2]    chr10         [60603, 60781]      * |  0.163240000605583
#>         [3]    chr10         [60782, 60816]      * |  0.372139990329742
#>         [4]    chr10         [60817, 60995]      * |  0.163240000605583
#>         [5]    chr10         [60996, 61625]      * | 0.0422799997031689
#>         ...      ...                    ...    ... .                ...
#>   [5789837]    chr10 [135524723, 135524734]      * |  0.144319996237755
#>   [5789838]    chr10 [135524735, 135524775]      * |  0.250230014324188
#>   [5789839]    chr10 [135524776, 135524784]      * |  0.427789986133575
#>   [5789840]    chr10 [135524785, 135524806]      * |  0.730019986629486
#>   [5789841]    chr10 [135524807, 135524837]      * |   1.03103005886078
#>   -------
#>   seqinfo: 25 sequences from hg19 genome
\end{verbatim}

@@

Returning to the Ranges object containing normalised coverage scores for
the methylation data, we can filter to find the coordinates of the peak
containing maximum coverage score. We can then find a 5000 nt region
centered around the maximum position.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{chr10_max_score_region <-}\StringTok{ }\NormalTok{chr10_scores }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{filter}\NormalTok{(score }\OperatorTok{==}\StringTok{ }\KeywordTok{max}\NormalTok{(score)) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{anchor_center}\NormalTok{() }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{set_width}\NormalTok{(}\DecValTok{5000}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

A common operation to perform between two \emph{Ranges} objects is to
find overlaps or nearest neighbours. The \emph{plyranges} API recasts
these operations as `joins' or `pairing' operations. For overlaps, there
are three join operations: \texttt{join\_overlap\_intersect()},
\texttt{join\_overlap\_inner()} and \texttt{join\_overlap\_left()} which
are shown in figure (). These operations consider any overlap between
two input ranges and return any corresponding metadata from both Ranges
objects as metadata. The intersect join takes the intersect of the start
and end coordinates of overlapping intervals of the query and subject
Ranges (for GenomicRanges it also accounts for sequence name), when
there is a overlap the metadata corresponding to the query and subject
Ranges are returned. Similarly, inner join takes the start and end
coordinates of the query Ranges that overlap the subject Ranges and
returns metadata of the overlapping query and subject Ranges. Finally,
the left join performs a left outer join between the query and subject
Ranges, it returns all genomic intervals from the query ranges, and
returns missing values in metadata columns when there is no overlap.

The overlap innner join could be used to restrict the chromosome 10
normalised coverage scores that are within the 5000nt region that
contains the max peak on chromosome 10 (visualised in figure ()).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{peak_region <-}\StringTok{ }\NormalTok{chr10_scores }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{join_overlap_inner}\NormalTok{(., }
\NormalTok{                     chr10_max_score_region }\OperatorTok{%>%}\StringTok{ }
\StringTok{                       }\KeywordTok{select}\NormalTok{(}\OperatorTok{-}\NormalTok{score))}
\NormalTok{ggbio}\OperatorTok{::}\KeywordTok{autoplot}\NormalTok{(peak_region, }
                \KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ score), }
                \DataTypeTok{geom =} \StringTok{"blank"}\NormalTok{) }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{geom_area}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ start))}
\end{Highlighting}
\end{Shaded}

\begin{figure}

{\centering \includegraphics{paper_files/figure-latex/peak_region-1}

}

\caption{Visualisation of peak coverage score accross 5000nt region of chromosome 10}\label{fig:peak_region}
\end{figure}

A user may also restrict or group by overlaps with the
\texttt{filter\_by\_overlaps()}, \texttt{filter\_by\_non\_overlaps()}
and \texttt{group\_by\_overlaps()}. All overlap methods can be modified
with the \texttt{within} suffix (which changes the type of overlap from
`any' to `within') or the \texttt{directed} suffix (which takes into
account the strand of a GenomicRanges object.).

@@
